---
title: "Introduction to Tidyverse"
author: "Zelvy Fauzan"
date: "2026-07-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is the second chapter of the DataCamp's Data Analyst in R career path.

The reason I wrote this article is to refresh and hone my skill in using R.

To be honest, I'm a career switcher from educational sciences to computer sciences. One year in and I learned 6 programming languages. I forgot all of them when I took an exam. :D 

Foolish, I know. But hey, a man's gotta survive, no?

---

## 1. Data Wrangling

*Data Wrangling* is when you are preparing the data that is going to be analyzed. It's the preparation step.

Stuffs like cleaning, filtering, arranging, ordering, mutating, etc.

We are going to use something called a `package`. Inside the package is the tools like scissor, hammer, wrench, and so on; but instead of that kind of tools, the package I'm talking about contains functions for cleaning, filtering, arranging, etc.

Say, you wanted to sort your data frame in R (or "table", if you aren't familiar with data frame terminology). R has its own function to sort data out, but the output is in vector.

Let's take the `gapminder` package and load the library and the dataset.

```{r dw1}
# install.packages("gapminder")
library(gapminder)
gapminder
```

Now let's see what will happen if we use the base R's `order()` function to our gapminder dataset. We want to order the country.

```{r dw2}
gapminder[order(gapminder$country),]
```

Well, that doesn't look *easy to read* for me. The code, I mean.

### 1.1. The dplyr Package

Luckily, we have a package called `dplyr`.

```{r dw3}
library(dplyr)
```

This package will help us with data management. Also, the functions that come with it is intuitive to read even for public audience. See for yourself below.

```{r dw4}
gapminder %>% 
  filter(year > 2000)
```

### 1.2. The Pipe Operator

The `%>%` is a pipe operator. It's equivalent to the word "then" when we are talking about procedures or step-by-step. Let me give you an example.

How to make an instant coffee:
1. Prepare the cup, instant coffee, water boiler.
2. Open the instant coffee and pour it into the cup.
3. Boil some water.
4. Wait until the water is boiled.
5. Pour the boiling water into the cup.

Now, if you are to give this instruction orally, you would say "then" after every step is done.

The `%>%` pipe operator is just like that.

### 1.3. The filter() function

The `filter()` function is self-explanatory. It filters the data frame based on the conditions that we specified. In this case, we want R to show us the data that has `year` larger than, say, `1999`.

```{r dw5}
gapminder %>% 
  filter(year > 1999)
```


### 1.4. The arrange() function

Next up is the `arrange()` function. Pretty much all of the functions in the `dplyr` package are straight forward. The verb is representative of its main function. We just need to adapt with the verb. 

For first timers, you may be more familiar with the word "sort" instead of "arrange". This is understandable because they have the same meaning. 

Linguistically, both verb have the same meaning too!

Let's take a look at how to arrange the data frame that we have here by the country name. From Z to A. Descending order.

(Because Z is the 26th alphabet and A is the 1st, so it's called descending... From biggest number to lowest number)

```{r dw6}
gapminder %>% 
  arrange(desc(country))
```


Compared to the `order()` function, this one is pretty easy to read, right?

And what if we want to order the data with multiple columns?

Well, say no more! The `dplyr` package got you covered.

```{r dw7}
gapminder %>% 
  arrange(desc(country), year, desc(lifeExp))
```

The code above *arranges* the data by the country in descending order.

Then, the result of that is arranged again by year ascendingly.

And finally, it arranges the data by lifeExp column in descending order.

Pretty easy, innit?

### 1.5. The mutate() function

The word *mutate* means to change something. This something can be new or replacing the existing one. I'm talking about columns here. 

If you take a look at the lifeExp of the gapminder dataset, you will see that it is in year unit. This means that if the value of lifeExp is 56.159 then this means that the average life expectation in that country in the given year is around 56.159 years. 

Let's assume we want to change the lifeExp column from year unit to month unit. We can do it by multiplying it with 12 (because 1 year is 12 months).

```{r dw8}
gapminder %>% 
  mutate(lifeExp = lifeExp * 12)
```

Ta-da!

It's in month unit now.

We can also make a new one instead of replacing the existing one. See the code below.

```{r dw9}
gapminder %>% 
  mutate(lifeExp_month = lifeExp * 12)
```

See? Easy peasy, lemon squishy.

---

## 2. Data Visualization

Now comes the exciting part. Data Visualization.

People don't see tables. People don't see matrix/matrices. Heck, they don't even see a vector. I'm saying this with the context of stakeholders. 

What people see and understand is graphics. In fact, human can easily understand graphical information rather than raw data.

Therefore, it is our job as a data analyst or data scientist to tell or communicate our findings. Let's start with some of the most basic graphics.

Oh, by the way, we are going to use the infamous `ggplot2` package.

```{r dv1}
# install.packages(ggplot2)
library(ggplot2)
```

First, let's say we are interested in seeing the `lifeExp` of Germany from time to time.

*Time to time* means from the earliest to the latest record available (i.e., 1952-2007).

```{r dv2}
gapminder %>% 
  filter(country == "Germany") %>% 
  arrange(year) %>% 
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line()
```

Pretty graphics!

This is sufficient but bare minimum by standard. It does not meet the standard of data communication. We will talk about it later on.

The function `ggplot()` has many arguments.

See, it has `aes()` function which stands for *aesthetics*. It means, we are telling the `ggplot()` to use whatever we put in the aes() as the basic properties of the chart.

The `aes()` function is also the place where we tell the ggplot what our variable of interests are. See the `x` and `y` arguments? That are our variable of interest.

You can specify different variables (columns) if you wanted to. Simply change the value of the `x` and `y` arguments and you are all set.

Let's create another graphics (or "chart", if you will). I'm thinking of looking at the continent level this time, and see the `lifeExp` variable over time.

```{r dv3}
gapminder %>% 
  filter(continent == "Europe") %>% 
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line()
```

Oh, no! It's a strange chart we get. I wonder what's wrong with this chart. Take a guess before continuing. If you get it right, treat yourself. :D

Since the variable continent is a group, not a single entity/country (or whatever you call it), this happened. It would be different if we use different `y` variable.

Let's do some experiments. I will change the `y` variable to country, and see what we get.

```{r dv4}
gapminder %>% 
  filter(continent == "Europe") %>% 
  ggplot(aes(x = year, y = country)) +
  geom_line()
```

Oh, no! Another strange chart that we get.

Take another guest of what I was about to show you from our last two charts. Again, if you guessed it right, treat yourself!

I was thinking of showing you the `lifeExp` of each country in the Europe continent. I wanted it to be a line chart. Each country is represented with a different color of line.

Take your time and save you answer. I will show you the demonstration of this problem in the next section.

For now, let's just move on.

### 2.1. The Bare Minimum of a Chart

I will show you the bare minimum of making a chart using ggplot2.

```{r}
gapminder %>% 
  filter(country == "Germany") %>% 
  ggplot(aes(year, lifeExp)) +
  geom_line() +
  labs(
    title = "Life Expectancy in Germany is Increasing Over the Year",
    subtitle = "This growth may also indicate improved economy of the country",
    caption = "Data was fetched from the 'Gapminder' dataset.",
    x = "Year",
    y = "Life Expectancy (year)"
  )
```

The `labs()` function inside the ggplot is for telling the R about the labels. 

* *title* is for the title.
* *subtitle* is for the subtitle (the text below the title).
* *caption* is the footnote, at the right bottom of the chart. We usually tell important sidenote about the chart like the source of the data or something.
* *x* is for the name of the variable on the x axis.
* *y* same as the x but this is for the y axis.

### 2.2. Log scales

There will be times where your data is too sparse. A small portion of the data are too high while the rest are small, or vice versa. Take a look below.

```{r dv5}
gapminder %>% 
  ggplot(aes(pop, gdpPercap)) +
  geom_point()
```

Most of the data points are gathered on the left bottom of the chart area. We can't see the bigger picture! We can't see what's happening there.

We need a way. We need to change the chart so that we can see the pattern.

Chart is made to show patterns, after all.

This is where log scale take the stage.

We can use a function that transform the scale from raw values to it's natural number (called the "log" or "natural log"). The function in called: `scale_x_log10()` and we call it after the `geom_point()`. Let's see the magic, shall we?

```{r dv6}
gapminder %>% 
  ggplot(aes(pop, gdpPercap)) +
  geom_point() +
  scale_x_log10()
```

Much better!

But as we see it, the data were rather fond of each other. They don't want to be separated from the other data points, it seems. Let's separate them using (you guessed it!) the `scale_y_log10()` function.

```{r dv7}
gapminder %>% 
  ggplot(aes(pop, gdpPercap)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

### 2.3. More than 2 Dimensions?



### 2.4. Facets: The More, The Merrier, no?





---

## 3. Grouping and Summarizing

Let's put the charts aside for a moment and shall we talk about summary?

### 3.1. Summarize

### 3.2. Grouping

### 3.3. Visualizing Summarized Data

`expand_limits(x, y)`

---

## 4. Types of Visualizations

### 4.1. Line plots

### 4.2. Bar plots

`geom_col()`: Vertical bars, as in a column which has values below it creating a vertically aligned data points.

`geom_bar()` : The opposite of `geom_col()` in that "bar" is horizontal. Equivalent to the *rows*.

### 4.3. Histograms

### 4.4. Boxplots

---

## Closing

Thank you for taking the time to read my article. I hope you gain something new or get some refreshment of the related topic.

---

If you are interested in reading more of my articles, check out [my first article on this article series](#).










